---
title: "Lab 9"
author: "Carissa Feliciano"
format: html
embed-resources: TRUE
---

# Problem 1: Vectorization

## Function 1 (n x k dataset)
```{r}

fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n){
    x <- rbind(x, rpois(k, lambda))    
  }
  
  return(x)
}

# Write a faster version of the function 
fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
}
```

```{r}
# Show that fun1alt generates a matrix with the same dimensions as fun1 
fun_matrix <- fun1()
dim(fun_matrix)

fun1alt_matrix <- fun1alt()
dim(fun1alt_matrix)
```

```{r}
# Show the values inside the two matrices follow similar distributions 

hist(as.vector(fun_matrix),
main = "Histogram of Values in the Fun1 Matrix",
xlab = "Value",
ylab = "Frequency")

hist(as.vector(fun1alt_matrix),
main = "Histogram of Values in the Fun1Alt Matrix",
xlab = "Value",
ylab = "Frequency")
```
The values inside the two matrices follow similar distributions. Both distributions are positively skewed. Both distributions have a peak around 3-4. 

```{r}
# Check the speed of the two functions 
microbenchmark::microbenchmark(
  fun1(),
  fun1alt()
)
```
The fun1alt function is approximately 9x faster than the fun1 function. The mean speed of the fun1 function is 211 microseconds. The mean speed of the fun1alt function is 39 microseconds. 

## Function 2 (max value of each column of a matrix)
```{r}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  x[cbind(max.col(t(x)), 1:ncol(x))]
}
```

```{r}
# Show that both functions return the same output for a given input matrix x

fun2_matrix <- fun2(x)
print(fun2_matrix)
```

```{r}
fun2alt_matrix <- fun2alt(x)
print(fun2alt_matrix)
```

```{r}
# Check the speed of the two functions
microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x)
)
```

The fun2alt function is 6.6x faster than the fun2 function.

# Problem 3: Parallelization
# 1. Parallelize the lapply loop
```{r}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
  
  #Parallelized function
  library(parallel)
  cl <- makeCluster(ncpus)
  clusterExport(cl, varlist = c("dat", "idx", "stat"), envir = environment())
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[, i], , drop = FALSE])
  })
  stopCluster(cl)
  
  # Converting the list into a matrix
  ans <- do.call(rbind, ans)

  return(ans)
}
```

# 2. Once you have a version of the my_boot() function that runs on multiple cores, check that it provides accurate results by comparing it to a parametric model:
```{r}
# Bootstrap of an OLS
my_stat <- function(d) coef(lm(y ~ x, data=d))

# DATA SIM
set.seed(1)
n <- 500; R <- 1e4

x <- cbind(rnorm(n)); y <- x*5 + rnorm(n)

# Checking if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)
```
# Check whether your version actually goes faster when it’s run on multiple cores
```{r}
system.time(my_boot(dat = data.frame(x, y), my_stat, R = 4000, ncpus = 1L))
system.time(my_boot(dat = data.frame(x, y), my_stat, R = 4000, ncpus = 2L))
```
It was faster to run it with multiple cores. 
